---
title: "Project 2"
format: html
author: "Group Members: Zoe & Owen"
---

```{r setup}
#| message: false
#| warning: false
#| include: false

library(ggplot2)
library(tidyverse)
library(mdsr)
library(DBI)
library(RMariaDB)
library(dbplyr)
library(duckdb)
library(readr)
```

### ---------------------------------------------------------------------------

# Accessing Data

Data accessing done by Zoe

```{r}
# Read in the three movie datasets from CSV files

# Dataset 1: Core movie info (genre, director, budget, gross, etc.)
movies <- readr::read_csv("movies.csv")

# Dataset 2: IMDb and Metascore ratings per movie
ratings <- readr::read_csv("movie_ratings.csv")

# Dataset 3: TMDB metadata (popularity, revenue, budget, release date)
tmdb <- readr::read_csv("tmdb_5000_movies.csv")
```

### ---------------------------------------------------------------------------

# Exploring the Data

Exploration done by Zoe

```{r}
# Explore the movies dataset - check variable types, column names, and missing values
glimpse(movies)
summary(movies)
```

```{r}
# Explore the ratings dataset
glimpse(ratings)
summary(ratings)
```

```{r}
# Explore the tmdb dataset
glimpse(tmdb)
summary(tmdb)
```

### ---------------------------------------------------------------------------

# Designing the Database

Database design done together by Zoe and Owen

```{r}
# Create a new empty DuckDB database called movie_db
# con is our connection object - we use it for all future SQL operations
con <- DBI::dbConnect(duckdb::duckdb(), dbdir = "movie_db")
```

```{r}
# Save each dataset as a CSV so we can load them using COPY FROM
# This follows the proper SQL workflow for importing data
write_csv(movies, "movies_clean.csv")
write_csv(ratings, "ratings_clean.csv")
write_csv(tmdb, "tmdb_clean.csv")
```

```{sql connection=con}
DROP TABLE IF EXISTS movies;
```

```{sql connection=con}
-- Create the movies table with explicit field types for each column
-- VARCHAR for text fields, INTEGER for whole numbers, DOUBLE for decimals
CREATE TABLE movies (
  title       VARCHAR(255) NOT NULL DEFAULT '',
  rating      VARCHAR(10) DEFAULT NULL,
  genre       VARCHAR(50) DEFAULT NULL,
  year        INTEGER DEFAULT NULL,
  released    VARCHAR(50) DEFAULT NULL,
  score       DOUBLE DEFAULT NULL,
  votes       INTEGER DEFAULT NULL,
  director    VARCHAR(255) DEFAULT NULL,
  writer      VARCHAR(255) DEFAULT NULL,
  star        VARCHAR(255) DEFAULT NULL,
  country     VARCHAR(100) DEFAULT NULL,
  budget      DOUBLE DEFAULT NULL,
  gross       DOUBLE DEFAULT NULL,
  company     VARCHAR(255) DEFAULT NULL,
  runtime     DOUBLE DEFAULT NULL
);
```

```{sql connection=con}
COPY movies FROM 'movies_clean.csv' (FORMAT CSV, HEADER, NULL 'NA', QUOTE '"', ESCAPE '"');
```

```{sql connection=con}
DESCRIBE movies;
```

```{sql connection=con}
SELECT * FROM movies LIMIT 10;
```

```{sql connection=con}
DROP TABLE IF EXISTS movie_ratings;
```

```{sql connection=con}
-- Create the movie_ratings table
-- imdb and metascore are decimals, votes is a whole number
CREATE TABLE movie_ratings (
  idx         INTEGER DEFAULT NULL,
  title       VARCHAR(255) NOT NULL DEFAULT '',
  year        INTEGER DEFAULT NULL,
  imdb        DOUBLE DEFAULT NULL,
  metascore   DOUBLE DEFAULT NULL,
  votes       INTEGER DEFAULT NULL
);
```

```{sql connection=con}
COPY movie_ratings FROM 'ratings_clean.csv' (FORMAT CSV, HEADER, NULL 'NA');
```

```{sql connection=con}
DESCRIBE movie_ratings;
```

```{sql connection=con}
SELECT * FROM movie_ratings LIMIT 10;
```

```{sql connection=con}
DROP TABLE IF EXISTS tmdb;
```

```{sql connection=con}
-- Load the full tmdb CSV directly, then select only the columns we need
-- The tmdb dataset has 20 columns; we extract the relevant ones for analysis
CREATE TABLE tmdb AS
SELECT title,
       CAST(release_date AS DATE) AS release_date,
       CAST(YEAR(CAST(release_date AS DATE)) AS INTEGER) AS year,
       CAST(budget AS DOUBLE) AS budget,
       CAST(revenue AS DOUBLE) AS revenue,
       CAST(popularity AS DOUBLE) AS popularity,
       CAST(vote_average AS DOUBLE) AS vote_average,
       CAST(vote_count AS INTEGER) AS vote_count,
       original_language,
       CAST(runtime AS DOUBLE) AS runtime,
       status
FROM read_csv_auto('tmdb_5000_movies.csv');
```

```{sql connection=con}
DESCRIBE tmdb;
```

```{sql connection=con}
SELECT * FROM tmdb LIMIT 10;
```

```{sql connection=con}
SHOW TABLES;
```

### ---------------------------------------------------------------------------

# Querying Data Using SQL

## Query 1 - Top 10 Most Profitable Movies

Done by Zoe — Overall goal: Which movies and genres are the most successful?

```{sql connection=con}
-- Calculate profit as revenue minus budget
-- Filter out rows where budget or revenue is missing (0)
SELECT title,
       year,
       budget,
       revenue,
       revenue - budget AS profit
FROM tmdb
WHERE revenue > 0 AND budget > 0
ORDER BY profit DESC
LIMIT 10
```

## Query 2 - Average IMDb Rating and Metascore by Genre

Done by Zoe

```{sql connection=con}
-- Join movies and movie_ratings on title and year to match films correctly
-- Group by genre to get average ratings per genre
SELECT m.genre,
       ROUND(AVG(r.imdb), 2) AS avg_imdb,
       ROUND(AVG(r.metascore), 2) AS avg_metascore,
       COUNT(*) AS movie_count
FROM movie_ratings r
JOIN movies m ON r.title = m.title
             AND r.year = m.year
GROUP BY m.genre
ORDER BY avg_imdb DESC
```

## Query 3 - Average Popularity and Revenue by Decade

Done by Zoe

```{sql connection=con}
-- Group movies into decades using integer division (e.g., 1987 / 10 * 10 = 1980)
-- Filter out rows with missing revenue
SELECT (year / 10) * 10 AS decade,
       ROUND(AVG(popularity), 2) AS avg_popularity,
       ROUND(AVG(revenue), 2) AS avg_revenue,
       COUNT(*) AS movie_count
FROM tmdb
WHERE year IS NOT NULL
  AND revenue > 0
GROUP BY decade
ORDER BY decade
```

## Query 4 - Full 3-Table JOIN

Done by Zoe

```{sql connection=con}
-- Join all three tables to combine genre/director info with ratings and TMDB metrics
-- Filter to only movies with known budgets and revenues
SELECT m.title,
       m.genre,
       m.director,
       r.imdb,
       r.metascore,
       t.budget,
       t.revenue,
       t.popularity
FROM movies m
JOIN movie_ratings r ON r.title = m.title
                    AND r.year = m.year
JOIN tmdb t ON t.title = m.title
           AND t.year = m.year
WHERE t.budget > 0 AND t.revenue > 0
ORDER BY t.popularity DESC
LIMIT 20
```

## Owen's Queries - Movie Ratings and Genre Trends Over Time

```{sql connection=con}
-- Join tmdb and movies to combine release year, content rating, genre, and vote average
-- Filter to only the four major content ratings for clarity
CREATE OR REPLACE TABLE table1 AS
SELECT EXTRACT(YEAR FROM CAST(release_date AS DATE)) AS year,
       FLOOR(EXTRACT(YEAR FROM CAST(release_date AS DATE)) / 10) * 10 AS decade,
       tmdb.title,
       movies.rating,
       movies.genre,
       vote_average
FROM tmdb
JOIN movies ON tmdb.title = movies.title
WHERE movies.rating IN ('R', 'PG', 'PG-13', 'G')
ORDER BY year
```

```{sql connection=con}
-- Count movies per year per rating and compute average votes
-- Used ChatGPT assistance to figure out the CROSS JOIN and proportion logic
CREATE OR REPLACE TABLE table2 AS
WITH base AS (
    SELECT year, rating,
           COUNT(*) AS count,
           AVG(vote_average) AS avg_votes
    FROM table1
    GROUP BY year, rating
),
years AS (SELECT DISTINCT year FROM table1),
ratings AS (SELECT DISTINCT rating FROM table1),
all_combos AS (
    SELECT y.year, r.rating
    FROM years y CROSS JOIN ratings r
)
SELECT a.year, a.rating,
       COALESCE(b.count, 0) AS count,
       COALESCE(b.count, 0) * 1.0
           / SUM(COALESCE(b.count, 0)) OVER (PARTITION BY a.year) AS proportion,
       b.avg_votes
FROM all_combos a
LEFT JOIN base b ON a.year = b.year AND a.rating = b.rating
WHERE count != 0
ORDER BY a.year
```

```{sql connection=con}
-- Same structure as table2 but grouped by genre instead of rating
CREATE OR REPLACE TABLE table3 AS
WITH base AS (
    SELECT year, genre,
           COUNT(*) AS count,
           AVG(vote_average) AS avg_votes
    FROM table1
    GROUP BY year, genre
),
years AS (SELECT DISTINCT year FROM table1),
genres AS (SELECT DISTINCT genre FROM table1),
all_combos AS (
    SELECT y.year, g.genre
    FROM years y CROSS JOIN genres g
)
SELECT a.year, a.genre,
       COALESCE(b.count, 0) AS count,
       COALESCE(b.count, 0) * 1.0
           / SUM(COALESCE(b.count, 0)) OVER (PARTITION BY a.year) AS proportion,
       b.avg_votes
FROM all_combos a
LEFT JOIN base b ON a.year = b.year AND a.genre = b.genre
WHERE count != 0
ORDER BY a.year
```

```{sql connection=con}
-- Combine table2 and table3 into a single summary table for visualization
CREATE OR REPLACE TABLE table4 AS
SELECT table2.year,
       table2.rating,
       table2.proportion AS rating_prop,
       table3.genre,
       table3.proportion AS genre_prop,
       AVG(table2.avg_votes) AS avg_votes
FROM table2
JOIN table3 ON table2.year = table3.year
GROUP BY table2.year, table2.rating, rating_prop, table3.genre, genre_prop
```

### ---------------------------------------------------------------------------

# Creating Data Visualizations

## Visualization 1 - Average IMDb Rating by Genre

Done by Zoe

```{r}
# Pull the genre ratings summary from the database into R for plotting
genre_ratings <- dbGetQuery(con, "
  SELECT m.genre,
         ROUND(AVG(r.imdb), 2) AS avg_imdb,
         COUNT(*) AS movie_count
  FROM movie_ratings r
  JOIN movies m ON r.title = m.title
               AND r.year = m.year
  GROUP BY m.genre
  ORDER BY avg_imdb DESC
")
```

```{r}
# Bar chart of average IMDb rating by genre
# reorder() sorts genres from lowest to highest avg_imdb for readability
# coord_flip() makes genre labels horizontal and easier to read
# scale_fill_gradient applies a blue color gradient - darker = higher rating
ggplot(genre_ratings, aes(x = reorder(genre, avg_imdb), y = avg_imdb, fill = avg_imdb)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(
    title = "Average IMDb Rating by Genre",
    subtitle = "Based on movies appearing in both datasets",
    x = "Genre",
    y = "Average IMDb Rating",
    fill = "Avg Rating"
  ) +
  theme_minimal()
```

## Visualizations 2, 3, 4 - Movie Trends Over Time (1980-2016)

Done by Owen

```{r}
# Pull table1 and table4 from the database into R for plotting
table1 <- DBI::dbGetQuery(con, "SELECT * FROM table1")
table4 <- DBI::dbGetQuery(con, "SELECT * FROM table4")

# Filter to only include movies from 1980 onwards
revised_table1 <- table1 |> filter(year >= 1980)
revised_table4 <- table4 |> filter(year >= 1980)

# Compute the average IMDb vote score per year across all movies
avg_votes_by_year <- revised_table1 |>
  group_by(year) |>
  summarise(avg_votes = mean(vote_average, na.rm = TRUE))
```

```{r}
# Line + point plot of average IMDb vote score by year
# The peach background (FFE5B4) is a consistent theme across Owen's visuals
ggplot(avg_votes_by_year, aes(x = year, y = avg_votes)) +
  geom_point(size = 3) +
  geom_line() +
  labs(x = "Year", y = "Average IMDB Votes",
       title = "Movie Popularity over Time (1980-2016)") +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(size = 10),
    legend.title = element_text(face = "bold"),
    legend.background = element_rect(fill = "white", color = "black"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    panel.grid.minor = element_blank(),
    plot.background = element_rect(fill = "#FFE5B4", color = NA)
  )
```

```{r}
# Stacked bar chart showing movies of each content rating released per year
ggplot(revised_table1, aes(x = year, fill = rating)) +
  geom_bar() +
  labs(x = "Year", y = "Count", fill = "Rating",
       title = "The Surge of Movie Rating Frequencies (1980-2016)") +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(size = 10),
    legend.title = element_text(face = "bold"),
    legend.background = element_rect(fill = "white", color = "black"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    panel.grid.minor = element_blank(),
    plot.background = element_rect(fill = "#FFE5B4", color = NA)
  )
```

```{r}
# Line chart showing the proportion of each genre per year
# Used ChatGPT assistance for theme and visual styling ideas
ggplot(revised_table4, aes(x = year, y = genre_prop, color = genre)) +
  geom_line() +
  labs(x = "Year", y = "Proportion", color = "Genre",
       title = "The Shift in Movie Genre Proportion (1980-2016)") +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(size = 10),
    legend.title = element_text(face = "bold"),
    legend.background = element_rect(fill = "white", color = "black"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    panel.grid.minor = element_blank(),
    plot.background = element_rect(fill = "#FFE5B4", color = NA)
  )
```

### ---------------------------------------------------------------------------

# Sources

-   <https://www.kaggle.com/datasets/danielgrijalvas/movies>
-   <https://www.kaggle.com/datasets/raidevesh05/movie-ratings-dataset>
-   <https://www.kaggle.com/datasets/tmdb/tmdb-movie-metadata>

### ---------------------------------------------------------------------------

```{r}
# Always disconnect from the database when done to free up resources
DBI::dbDisconnect(con, shutdown = TRUE)
```
